---
title: "GH_YF_Incidence_LISA"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(pacman)

p_load(sf, tidyverse, ggplot2, ggspatial, spdep)
```

## Import shapefile and annual confirmed/probable case data

```{r}
gh_sf <- st_read("../Shapefiles/261_Districts.shp") %>% 
  st_make_valid()


# remove row with no values in geometry 
gh_sf <- gh_sf %>%
  filter(!is.na(OBJECTID))

load("annualYF_1970_2024.RData")

annual_dist <- read.csv("confirmed_probable_2005_2024.csv")
```


# Import population data and calculate incidence rates
```{r}
pop_5yr <- read.csv("district_populations_2000_2025.csv") %>% 
  filter(year < 2025 & year > 2004) %>% 
  dplyr::select(-X)


five_yr_cases_dist <- annual_dist %>% 
  left_join(pop_5yr, by = c("district", "year")) %>% 
  mutate(range = case_when(
    year %in% 2005:2009 ~ "2005-2009",
    year %in% 2010:2014 ~ "2010-2014",
    year %in% 2015:2019 ~ "2015-2019",
    year %in% 2020:2024 ~ "2020-2024",
  )) %>% 
  group_by(district, range) %>% 
  summarise(tot_cases = sum(cases),
            tot_pop = sum(pop)) %>% 
  arrange(district)

sum(five_yr_cases_dist$tot_cases)

five_district <- five_yr_cases_dist %>% 
  # incidence per 100,000 person-years
  mutate(incidence = ((tot_cases / tot_pop)*100000))

```

# Map incidence rates for confirmed/probable cases
```{r}
cases5yr_sf <- gh_sf %>%
  left_join(five_district, by = c("District" = "district")) %>% 
  filter(!is.na(incidence))

bbox <- st_bbox(gh_sf)

scale_arrow_df <- st_as_sf(data.frame(
  range = "2005-2009",
  x = bbox["xmin"] - 0.1,   # right side
  y = bbox["ymin"] + 0.1    # BOTTOM
), coords = c("x", "y"), crs = st_crs(gh_sf))



gh_5yr_IR <- ggplot() +
  geom_sf(data = gh_sf, fill = "white", color = "darkgrey") +
  geom_sf(data = cases5yr_sf, aes(fill = incidence), color = "darkgrey") +
  scale_fill_viridis_c(
    na.value = "grey90",
    name = "Probable/Confirmed YF Cases\nper 100,000 person-years"
  ) +
  facet_wrap(~range, nrow = 1) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  ) +
 annotation_scale(
    data = scale_arrow_df,
    location = "br",
    width_hint = 0.25,
    text_cex   = 0.55,
    line_width = 0.6,
    pad_x      = unit(0.2, "npc"),
    pad_y      = unit(0.02, "npc")
  ) +

  annotation_north_arrow(
    data = scale_arrow_df,
    location = "br",
    which_north = "true",
    height = unit(0.6, "cm"),
    width  = unit(0.6, "cm"),
    pad_x  = unit(0.02, "npc"),
    pad_y  = unit(0.02, "npc"),
    style = north_arrow_fancy_orienteering()
  )

gh_5yr_IR
```


```{r}
ggsave(
filename = file.path("Results/GH_ProbConf_Inc_maps.tiff"),
plot = gh_5yr_IR,
device = "tiff",
dpi = 300,
width = 200,
height = 140,
units = "mm",
compression = "lzw"
)
```

# Import suspected annual case data and calculate incidence rates
```{r}
clean_district <- function(x){
  x %>%
    str_to_lower() %>%
    str_replace_all("-", " ") %>%
    str_replace_all("[^a-z ]", "") %>%
    str_squish()
}

pop_5yr$district_final <- clean_district(pop_5yr$district)


annual_sus <- read.csv("suspected_YF_2005_2024.csv")


five_yr_sus_dist <- annual_sus %>% 
  filter(year > 2004) %>% 
  left_join(pop_5yr, by = c("district_final", "year")) %>% 
  mutate(range = case_when(
    year %in% 2005:2009 ~ "2005-2009",
    year %in% 2010:2014 ~ "2010-2014",
    year %in% 2015:2019 ~ "2015-2019",
    year %in% 2020:2024 ~ "2020-2024",
  )) %>% 
  group_by(district_final, range) %>% 
  summarise(tot_cases = sum(cases),
            tot_pop = sum(pop)) %>% 
  arrange(district_final)

sum(five_yr_sus_dist$tot_cases)

five_district_sus <- five_yr_sus_dist %>% 
  # incidence per 100,000 person-years
  mutate(incidence = ((tot_cases / tot_pop)*100000))
```

# Map incidence rates for suspected cases
```{r}
gh_sf$district_final <- clean_district(gh_sf$District)

cases5yr_sus_sf <- gh_sf %>%
  left_join(five_district_sus, by = "district_final")  %>% 
  filter(!is.na(incidence))


gh_5yr_sus_IR <- ggplot() +
  geom_sf(data = gh_sf, fill = "white", color = "darkgrey") +
  geom_sf(data = cases5yr_sus_sf, aes(fill = incidence), color = "darkgrey") +
  scale_fill_viridis_c(
    na.value = "grey90",
    name = "Suspected YF Cases\nper 100,000 person-years"
  ) +
  facet_wrap(~range, nrow = 1) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  ) +
 annotation_scale(
    data = scale_arrow_df,
    location = "br",
    width_hint = 0.25,
    text_cex   = 0.55,
    line_width = 0.6,
    pad_x      = unit(0.2, "npc"),
    pad_y      = unit(0.02, "npc")
  ) +
  annotation_north_arrow(
    data = scale_arrow_df,
    location = "br",
    which_north = "true",
    height = unit(0.6, "cm"),
    width  = unit(0.6, "cm"),
    pad_x  = unit(0.02, "npc"),
    pad_y  = unit(0.02, "npc"),
    style = north_arrow_fancy_orienteering()
  )

gh_5yr_sus_IR

```


```{r}
ggsave(
filename = file.path("Results/GH_Susp_Inc_maps.tiff"),
plot = gh_5yr_sus_IR,
device = "tiff",
dpi = 300,
width = 200,
height = 140,
units = "mm",
compression = "lzw"
)
```


## LISA 

# Confirmed/probable cases
```{r}
# expand dataset to include all districts, including those with zero cases
ranges <- sort(unique(five_district$range))

# dataset with one row for each district in each timeframe
sf_poly_ranges <- gh_sf %>%
  slice(rep(1:n(), each = length(ranges))) %>%
  mutate(range = rep(ranges, times = nrow(gh_sf)))

morans_confirmed <- sf_poly_ranges %>%
  left_join(five_district %>% dplyr::select(district, range, incidence),
            by = c("District" = "district", "range")) %>%
  mutate(incidence_5yr = replace_na(incidence, 0))

```

```{r}
# Create neighbors object
gh_sp <- as(gh_sf, "Spatial")
neighbors <- poly2nb(gh_sp, queen = TRUE)
weights <- nb2listw(neighbors, style = "W")
```

```{r}
local_morans_5yrs_calc <- function(data_df, year_col, inc_col, lw) {
  
  years <- sort(unique(data_df[[year_col]]))
  
  results <- lapply(years, function(y) {
    
    inc_y <- data_df %>%
      filter(.data[[year_col]] == y) %>%
      pull(.data[[inc_col]])
    
    local_moran <- localmoran(inc_y, lw, na.action = na.exclude)
    
    tibble(
      range = y,
      District = data_df %>% 
        filter(.data[[year_col]] == y) %>% 
        pull(District),
      local_I = local_moran[, "Ii"],
      z_I = local_moran[, "Z.Ii"],
      p_value = local_moran[, 5],
      cluster = attr(local_moran, "quadr")
    )
  })
  
  bind_rows(results)
}

local_morans_5yrs <- local_morans_5yrs_calc(
  data_df = morans_confirmed,
  year_col = "range",
  inc_col = "incidence_5yr",
  lw = weights
)


print(local_morans_5yrs)
```

# Map LISA of confirmed/probable cases
```{r}
sf_local_moran <- sf_poly_ranges %>%
  left_join(local_morans_5yrs,
            by = c("District", "range"))

class(sf_local_moran)


local_moran_map <- ggplot(sf_local_moran) +
  geom_sf(aes(fill = cluster$median), color = "white", size = 0.1) +

  scale_fill_manual(
    values = c(
      "High-High" = "#d73027",
      "Low-Low" = "#4575b4",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9"
    ),
    name = "LISA Cluster Type"
  ) +
  facet_wrap(~ range, nrow = 1) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  ) +
 annotation_scale(
    data = scale_arrow_df,
    location = "br",
    width_hint = 0.25,
    text_cex   = 0.55,
    line_width = 0.6,
    pad_x      = unit(0.2, "npc"),
    pad_y      = unit(0.02, "npc")
  ) +
  annotation_north_arrow(
    data = scale_arrow_df,
    location = "br",
    which_north = "true",
    height = unit(0.6, "cm"),
    width  = unit(0.6, "cm"),
    pad_x  = unit(0.02, "npc"),
    pad_y  = unit(0.02, "npc"),
    style = north_arrow_fancy_orienteering()
  )


local_moran_map
```

```{r}
ggsave(
filename = file.path("Results/GH_local_morans_confirmed_probable.tiff"),
plot = local_moran_map,
device = "tiff",
dpi = 300,
width = 200,
height = 140,
units = "mm",
compression = "lzw"
)
```

# LISA supspected cases
```{r}

local_morans_5yrs_sus_calc <- function(data_df, year_col, inc_col, lw) {
  
  years <- sort(unique(data_df[[year_col]]))
  
  results <- lapply(years, function(y) {
    
    inc_y <- data_df %>%
      filter(.data[[year_col]] == y) %>%
      pull(.data[[inc_col]])
    
    local_moran <- localmoran(inc_y, lw, na.action = na.exclude)
    
    tibble(
      range = y,
      district_final = data_df %>% 
        filter(.data[[year_col]] == y) %>% 
        pull(district_final),
      local_I = local_moran[, "Ii"],
      z_I = local_moran[, "Z.Ii"],
      p_value = local_moran[, 5],
      cluster = attr(local_moran, "quadr")
    )
  })
  
  bind_rows(results)
}

local_morans_5yrs_sus <- local_morans_5yrs_sus_calc(
  data_df = morans_sus,
  year_col = "range",
  inc_col = "incidence_5yr",
  lw = weights
)

sf_local_moran_sus <- sf_poly_ranges %>%
  left_join(local_morans_5yrs_sus,
            by = c("district_final", "range"))


```

# Map LISA of suspected cases
```{r}
local_moran_map_sus <- ggplot(sf_local_moran_sus) +
  geom_sf(aes(fill = cluster$median), color = "white", size = 0.1) +

  scale_fill_manual(
    values = c(
      "High-High" = "#d73027",
      "Low-Low" = "#4575b4",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9"
    ),
    name = "LISA Cluster Type"
  ) +
  facet_wrap(~ range, nrow = 1) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  ) +
 annotation_scale(
    data = scale_arrow_df,
    location = "br",
    width_hint = 0.25,
    text_cex   = 0.55,
    line_width = 0.6,
    pad_x      = unit(0.2, "npc"),
    pad_y      = unit(0.02, "npc")
  ) +
  annotation_north_arrow(
    data = scale_arrow_df,
    location = "br",
    which_north = "true",
    height = unit(0.6, "cm"),
    width  = unit(0.6, "cm"),
    pad_x  = unit(0.02, "npc"),
    pad_y  = unit(0.02, "npc"),
    style = north_arrow_fancy_orienteering()
  )


local_moran_map_sus
```


```{r}
ggsave(
filename = file.path("Results/GH_local_morans_suspected.tiff"),
plot = local_moran_map_sus,
device = "tiff",
dpi = 300,
width = 200,
height = 140,
units = "mm",
compression = "lzw"
)
```

# Line graphs of incidence rate by ecological zone
# incidence rate over full study period
```{r}
# create sequence of all study years
all_years <- seq(from = 2005, to = 2024, by = 1)

# make set of all years for each district
district_year_grid <- expand_grid(
  district = gh_sf$District,
  year = all_years
)

districts_final <- tibble(
  district = gh_sf$District,
  district_final = gh_sf$district_final
)

district_year_grid <- district_year_grid %>% 
  left_join(districts_final, by = "district")
  

# join cases to grid
cases_conf <- district_year_grid %>%
  left_join(annual_dist, by = c("district", "year")) %>%
  mutate(tot_cases = replace_na(cases, 0),
         type = "Confirmed/Probable") %>% 
  rename(cases_confirmed = cases)

cases_conf <- cases_conf %>% 
  left_join(zones, by = "district_final") %>% 
  left_join(pop_5yr, by = c("district_final", "year")) %>% 
  group_by(year, zone, type) %>% 
  summarise(tot_cases = sum(tot_cases),
            tot_pop = sum(pop)) %>% 
  mutate(incidence = ((tot_cases / tot_pop)*100000))

cases_sus <- district_year_grid %>%
  left_join(annual_sus, by = c("district_final", "year")) %>%
  mutate(tot_cases = replace_na(cases, 0),
         type = "Suspected") %>% 
  rename(cases_suspected = cases)

cases_sus <- cases_sus %>% 
  left_join(zones, by = "district_final") %>% 
  left_join(pop_5yr, by = c("district_final", "year")) %>% 
  group_by(year, zone, type) %>% 
  summarise(tot_cases = sum(tot_cases),
            tot_pop = sum(pop)) %>% 
  mutate(incidence = ((tot_cases / tot_pop)*100000))

cases_inc_tot <- bind_rows(cases_conf, cases_sus)

```

```{r}

inc_by_zone <- ggplot(cases_inc_tot, aes(x = year, y = incidence, color = zone)) +
  geom_line(linewidth = 1) +
  # geom_point() + 
  facet_wrap(~ type, scales = "free_y") +
  scale_color_manual(values = cols) +
  labs(x = "Year", y = "Incidence per 100k", color = "Ecological Zone") +
  theme_bw(base_size = 14) +
  theme(legend.position = "bottom")
  
inc_by_zone

ggsave(
filename = file.path("Results/inc_zone_year.tiff"),
plot = inc_by_zone,
device = "tiff",
dpi = 300,
width = 275,
height = 140,
units = "mm",
compression = "lzw"
)

```
