---
title: "GH_ENMeval_YF"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Models:
  - Initial covariates include:
    - WorldClim (excluding bio 8, 9, 18, 19 due to discontinuities); bio6 due to 
    response curve not aligning with a priori hypothesis
    - NDVI (avg. 2000-2024)
    - population density (avg 2000-2020, WorldPop 1km, UN adj)
    - Elevation
    - Built-up surface area (avg 2005-2025 in 5 year increments, 1km, GHS BUILT)
    - edge density (2019, continuous)
  - KD bias raster used for background point selection
  - 4 partitions made with hierarchical checkerboard
  - Includes data all from 2005-2024
  
  code adapted from https://jamiemkass.github.io/ENMeval/
  
## Load packages

```{r}
library(pacman)

p_load(geodata, 
       tidyverse, 
       terra, 
       sf, 
       ENMeval,
       dismo,
       ecospat,
       caret, 
       spatstat,
       sparr,
       tidyterra,
       ggplot2,
       ggspatial,
       thematic,
       patchwork,
       paletteer,
       gt
)
```

## Set seed for reproducibility

```{r}
set.seed(15) #set random seed in order to reproduce this analysis
```

## Import shapefile

```{r}
gh_sf <- st_read("Ghana_ENM_Covariates/261_Districts.shp") %>% 
  st_make_valid()


# remove row with no values in geometry 
gh_sf <- gh_sf %>%
  filter(!is.na(OBJECTID))

# Convert Ghana polygon to SpatVector
gh_vect <- terra::vect(gh_sf)
```

## Import mean annual NDVI - averaged from 2000-2020

```{r}
ndvi <- rast("Ghana_ENM_Covariates/Mean_NDVI_Ghana_2000_2024.tif")



# create template with extent of shapefile
template <- crop(ndvi, gh_vect)
template <- mask(template, gh_vect)

plot(template)

#writeRaster(template, "template_raster.tif", overwrite = TRUE)

plot(ndvi)

```


## Import WorldClim

```{r}
# Import data using geodata package
bio <- worldclim_country("GHA", # ghana
                        var = "bio", # all bioclimatic (19)
                        res = 0.5, # 0.5 resolution
                        path = "Ghana")


names(bio) <- c("annual_mean_temp", "mean_diurnal_range_temp", "isothermality",
                "temp_seasonality", "max_temp_warmest_month", "min_temp_coldest_month",
                "temp_annual_range", "mean_temp_wettest_quarter", "mean_temp_driest_quarter",
                "mean_temp_warmest_quarter", "mean_temp_coldest_quarter", "annual_precip",
                "precip_wettest_month", "precip_driest_month", "precip_seasonality",
                "precip_wettest_quarter", "precip_driest_quarter", "precip_warmest_quarter",
                "precip_coldest_quarter")


bio <- bio %>% 
  terra::project(template, method = "bilinear") %>% 
  terra::crop(template) %>% 
  terra::mask(template)

# elevation data
gh_elev <- rast("Ghana_ENM_covariates/gh_elevation_worldclim.tif") 
```




## Import population density - UN adjusted, 2020, 1km

source: https://data.apps.fao.org/catalog/dataset/304c21fb-0f5a-44ad-9948-2af6a7144fb5

```{r}
pop <- rast("Ghana_ENM_Covariates/gh_pd_05_20.tif") %>%
  terra::project(template, method = "bilinear") %>%
  terra::crop(template) %>%
  terra::mask(template)

plot(pop)

names(pop) <- "population_density"



```


## Import edge density raster - 2019
```{r}
ed <- rast("Ghana_ENM_Covariates/edge_density_window.tif") %>% 
  terra::project(template, method = "bilinear") %>% 
  terra::crop(template) %>% 
  terra::mask(template)

plot(ed)

names(ed) <- "edge_density"

```


## Import built-up surface area - averaged from 2005

```{r}
built <- rast("Ghana_ENM_Covariates/GHSL_Mean_BuiltSurface_05_25.tif") %>% 
  terra::project(template, method = "bilinear") %>% 
  terra::crop(template) %>% 
  terra::mask(template)

plot(built)
```

## Check covariance among covariates
```{r}
# Align to the template
ndvi_aligned <- terra::resample(ndvi, template, method = "bilinear")
pop_aligned  <- terra::resample(pop, template, method = "bilinear")
elev_aligned <- terra::resample(gh_elev, template, method = "bilinear")
built_aligned <- terra::resample(built, template, method = "bilinear")
ed_aligned <- terra::resample(ed, template, method = "bilinear")
bio_aligned <- terra::resample(bio, template, method = "bilinear")


# Stack all layers
envs <- c(bio_aligned, ndvi_aligned, elev_aligned, pop_aligned,
          built_aligned, ed_aligned)

names(envs)

# scale all covariates
envs <- scale(envs)

# Create a mask for cells valid in all layers
valid_cells <- app(envs, fun = function(...) all(!is.na(c(...))))

# remove bio8, bio9, bio18, bio19 (Booth, 2022, Judson et al., 2024)
envs <- envs[[names(envs) %in% c("annual_mean_temp", "mean_diurnal_range_temp", "isothermality",
                "temp_seasonality", "max_temp_warmest_month",
                "temp_annual_range", 
                "mean_temp_warmest_quarter", "mean_temp_coldest_quarter",
                "annual_precip", "precip_wettest_month", "precip_driest_month", 
                "precip_seasonality", "precip_wettest_quarter", "precip_driest_quarter",
                "NDVI_mean", "elevation", "built_surface", "edge_density")]]

# Apply mask to the whole stack
envs <- mask(envs, valid_cells, maskvalue = 0)

# Pull data as a data.frame
vals <- as.data.frame(envs, xy = FALSE)

# remove any NAs
vals <- na.omit(vals)

summary(vals) # checking for no NAs

# run correlation matrix
cor_mat <- cor(vals, use = "pairwise.complete.obs")

# ID highly correlated variables
to_remove <- findCorrelation(cor_mat, cutoff = 0.7, verbose = TRUE)

# Pull variables to keep 
vars_keep <- colnames(cor_mat)[-to_remove]
vars_keep

# subset raster
envs_filtered_pop <- envs[[vars_keep]]

# check remaining correlation coeff
cor_filtered <- cor(as.data.frame(envs_filtered_pop, xy = FALSE))
print(round(cor_filtered, 2))


# Open TIFF device
tiff(filename = "Final_Results/ENM_covariates.tif",
     width = 180, height = 120, units = "mm", res = 300, compression = "lzw")

plot(envs_filtered_pop)

dev.off()

plot(envs_filtered_pop)

```

## Assess if areas are climatically different if they are associated with occurrence points

Method: Multivariate Environmental Similarity surface, or MESS (Elith et al.2010)

Higher positive values indicated more similarity.

```{r}
occs_train <- read.csv("ENM_inputs_final/gh_occs_train_05_24.csv")

occs_test  <- read.csv("ENM_inputs_final/gh_occs_test_05_24.csv")


# extract climatic variables at occurrences and filter out occurrences with NAs for environmental variables
occs.z <- terra::extract(envs_filtered_pop, occs_train, ID = FALSE, xy = TRUE)

occs.na <- which(rowSums(is.na(occs.z)) > 0)

if (length(occs.na) > 0) { 
  occs <- occs[-occs.na, ] 
  occs.z <- occs.z[-occs.na, ] %>% dplyr::select(-c(x,y)) 
  occs_train <- occs_train[-occs.na, ]
} else { 
    # keep all rows; just drop x/y from occs.z 
  occs.z <- occs.z %>% dplyr::select(-c(x,y)) }


occs.mess <- predicts::mess(envs_filtered_pop, occs.z)

plot(occs.mess, main = "Environmental similarity")
points(occs_train)


# 97 training occurrences
# 33 test occurrences
```

## Create background sampling points using occurrence kernel desnity as bias raster

Create study window and ppp object

```{r}
# convert to meter projection
gh_m <- st_transform(gh_sf, crs = 32630)

occs_sf <- st_as_sf(occs_train, coords = c("x", "y"), crs = 4326)
occs_sf <- st_transform(occs_sf, crs = st_crs(gh_m))

st_bbox(occs_sf)    # points in UTM meters
st_bbox(gh_m)  # polygon in UTM meters

# create definition for the study region
gh_owin <- as.owin(gh_m)

# make a ppp object for case data
cases_ppp <- ppp(
  x = st_coordinates(occs_sf)[, 1],  # all X coordinates
  y = st_coordinates(occs_sf)[, 2],  # all Y coordinates
  window = gh_owin            # study area
)

```

Ripley’s L Plots with MC envelope 
```{r eval=FALSE}
# produce L-function with envelope
 # Envelope function
 #   lf.env <- envelope(cases_ppp, Lest, nsim = 99, correction="isotropic")
 # 
 #  # # Center it by subtracting r
 #   lf.env$obs <- lf.env$obs - lf.env$r
 #   lf.env$lo  <- lf.env$lo  - lf.env$r
 #   lf.env$hi  <- lf.env$hi  - lf.env$r
 #   lf.env$theo <- 0  # theoretical CSR line at 0
 # 
 #   plot(lf.env, main = NULL,
 #       legendargs = list(x = "topleft", inset = c(0.0, 0), cex = 0.7))
 # 
 #  # #Find peak distance as a candidate for bandwidth
 #  # # L(r) - r from observed curve
 #   L_diff <- lf.env$obs
 # 
 #  # # find distance of strongest clustering
 #  peak_idx <- which.max(L_diff)
 #   dist_peak <- lf.env$r[peak_idx]
 # 
 # dist_peak

 
 # 30155.51
```

```{r}
#  Kernel density (sigma is in meters in this CRS)
sigma_val <- 25000 # ~ peak r from Ripley's K/L

# increased dimensions to smooth edges in final KDE plot
dens <- density.ppp(cases_ppp, sigma = sigma_val, at = "pixels", edge = TRUE)

# convert to raster
kde <- rast(dens) 

crs(kde) <- "EPSG:32630"

kde <- kde %>% 
  terra::project(envs_filtered_pop, method = "bilinear") %>% 
  terra::crop(envs_filtered_pop) %>% 
  terra::mask(envs_filtered_pop)

kde <- kde[[1]]

plot(kde)
```


```{r}
# normalize
bias_raster <- kde / as.numeric(global(kde, "sum", na.rm = TRUE))

# should be 1
global(bias_raster, "sum", na.rm = TRUE)

# check if population density is correlated to bias correction
cor(values(pop),
    values(bias_raster),
    use="complete.obs") # 0.17

bg <- spatSample(
  bias_raster,
  size = 10000,
  method = "weights",
  na.rm = TRUE,
  xy = TRUE,
  values = FALSE
) %>% 
  as.data.frame()

colnames(bg) <- colnames(occs_train)

# check coverage of background points
plot(envs_filtered_pop[[1]])
points(bg, pch = 20, cex = 0.2)
```

## Partition using hierarchcial checkerboard 
```{r}
# Hierarchical checkerboard with k=4 spatial groups
cb2 <- get.checkerboard(occs_train, envs_filtered_pop, bg, aggregation.factor = c(8,8))
evalplot.grps(pts = occs_train, pts.grp = cb2$occs.grp, envs = envs_filtered_pop)

table(cb2$occs.grp)
table(cb2$bg.grp)

# Plotting the background points on checkerboard
evalplot.grps(pts = bg, pts.grp = cb2$bg.grp, envs = envs_filtered_pop)

# extract predictor variable values at background localities
bg.z <- cbind(bg, terra::extract(envs_filtered_pop, bg, ID = FALSE, xy = FALSE)) %>% 
  dplyr::select(-c(x,y))


# plot histograms or rasters of MESS predictions with each partition as the reference.
evalplot.envSim.hist(occs.z = occs.z, bg.z = bg.z, occs.grp = cb2$occs.grp, 
                     bg.grp = cb2$bg.grp, ref.data = "occs_train")

```

## Model with simple FC
```{r}
user.grp2 <- list(occs.grp = cb2$occs.grp, 
                 bg.grp = cb2$bg.grp)

m_kd_bias <- ENMevaluate(occs = occs_train, 
                   envs = envs_filtered_pop, 
                   bg = bg, 
                   algorithm = 'maxnet', 
                   partitions = "user",
                   user.grp = user.grp2,
                   tune.args = list(fc = c("L", "Q", "LQ"), 
                                    rm = c(0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5)))

```

```{r}
# Overall results
m_kd_bias.res <- eval.results(m_kd_bias)
m_kd_bias.res

```

```{r}
#select models with the highest average validation CBI, and to break ties, with the lowest average test omission rate
opt.seq_kd_bias <- m_kd_bias.res %>%
  filter(cbi.val.avg == max(cbi.val.avg)) %>%
  filter(or.10p.avg == min(or.10p.avg))

opt.seq_kd_bias 

opt.tune <- as.character(opt.seq_kd_bias$tune.arg)

# partition results
partitions <- eval.results.partitions(m_kd_bias) %>% 
  filter(tune.args == opt.tune)

partitions

metrics = opt.seq_kd_bias %>%
  mutate(auc.val_low =  auc.val.avg - auc.val.sd,
         auc.val_high = auc.val.avg + auc.val.sd,
         cbi.val_low = cbi.val.avg - cbi.val.sd,
         cbi.val_high = cbi.val.avg + cbi.val.sd,
         or10_low = or.10p.avg - or.10p.sd,
         or10_high = or.10p.avg + or.10p.sd) %>% 
  pivot_longer(cols = c(auc.val.avg, auc.val_low, auc.val_high,
                        cbi.val_high, cbi.val_low, cbi.val.avg,
                        or10_low, or10_high, or.10p.avg),
               names_to = "metric",
               values_to = "value") %>% 
  select(metric, value) 

metrics
```

## Optimal model response curves
```{r}
mod.seq_kd_bias <- eval.models(m_kd_bias)[[opt.seq_kd_bias$tune.args]] #select optimal model

mod.seq_kd_bias$betas # non-zero coefficients in model

# Open TIFF device
tiff(filename = "Final_Results/ENM_response_curves.tif",
     width = 180, height = 120, units = "mm", res = 300, compression = "lzw")

plot(mod.seq_kd_bias, type = "cloglog") #plot response curves

dev.off()

```

```{r}
pred.seq_kd_bias <- eval.predictions(m_kd_bias)[[as.character(opt.seq_kd_bias$tune.args)]] #prediction based on selected optimal model above
plot(pred.seq_kd_bias)
points(eval.occs(m_kd_bias), pch = 21, bg = eval.occs.grp(m_kd_bias))

writeRaster(pred.seq_kd_bias, "Final_Results/env_suitability_pred.tif", overwrite = TRUE)

```

```{r include = FALSE}
#pred_map <- rast("Results/env_suitability_pred.tif")
```

# Evaluate model on test points (n = 33)
```{r}
# create a vector of suitability predictions
pred_vec <- pred.seq_kd_bias[]
pred_vec <- pred_vec[is.finite(pred_vec)]   # drops NA, NaN, Inf

summary(pred_vec)

# range should be 0-1
range(pred_vec)

# extract suitability at test occurrence locations
coords_test <- as.matrix(occs_test[, c("x","y")])

test.df <- terra::extract(pred.seq_kd_bias, coords_test)
# pull numeric vector
test.vals <- test.df[,1]  


test_cbi <- ecospat.boyce(pred_vec, test.vals)
test_cbi$cor


# calculate test OR10
# extract suitability at training locations
coords_train <- as.matrix(occs_train[, c("x","y")])
train_vals <- terra::extract(pred.seq_kd_bias, coords_train)[,1]

# compute 10th percentile threshold
th_or10 <- quantile(train_vals, 0.10, na.rm = TRUE) #0.28
th_or10

# compute test data OR10
OR10_test <- mean(test.vals < th_or10)
OR10_test 


sens_OR10_test <- mean(test.vals >= th_or10)
sens_OR10_test 

plot(pred.seq_kd_bias, axes = FALSE, box = FALSE)
points(occs_test, pch=20, col="red")

occs_test$occ_label <- "Test Occurrence (2024)"


suitability_map_cont <- ggplot() +
  geom_spatraster(data = pred.seq_kd_bias) +
    # geom_point(
    # data = occs_test,
    # aes(x = x, y = y, color = occ_label)) +
  paletteer::scale_fill_paletteer_c("grDevices::Mako", name = "Environmental Suitability", na.value = "transparent",
                                    direction = -1) +
    theme_bw() + #Clean theme
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.ticks = element_blank(), # Remove axis ticks,
    axis.text = element_blank(),  # Remove axis text (coordinate degrees)
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank()
) +
  coord_sf(crs = "EPSG:4326", expand = FALSE) +
    annotation_scale(
    location   = "bl",
    width_hint = 0.25,
    text_cex   = 0.55,
    line_width = 0.6,
    pad_x      = unit(0.5, "npc"),
    pad_y      = unit(0.03, "npc")
  ) +
  annotation_north_arrow(
    location = "br",
    which_north = "true",
    height = unit(0.8, "cm"),
    width  = unit(0.8, "cm"),
    pad_x  = unit(0.02, "npc"),
    pad_y  = unit(0.02, "npc"),
  style = north_arrow_fancy_orienteering()
)

suitability_map_cont
```

```{r}
ggsave(
filename = file.path("Final_Results/Cont_Suitability.tiff"),
plot = suitability_map_cont,
device = "tiff",
dpi = 300,
width = 180,
height = 120,
units = "mm",
compression = "lzw"
)
```


# Save ENM inputs for SDMtune
```{r}
write.csv(bg, "ENM_inputs_final/gh_bg_05_24.csv")
writeRaster(envs_filtered_pop, "ENM_inputs_final/gh_envs_05_24.tif", overwrite = TRUE)
```

## MTP Threshold map
Source: https://github.com/jamiemkass/ENMeval/issues/53
```{r}
names(eval.predictions(m_kd_bias))

# Pull out the spatial predictions for best model
pred.r <- eval.predictions(m_kd_bias)[[opt.tune]]

# extract occ coordinates
occ.xy <- m_kd_bias@occs[, 1:2]

# Extract the predicted values for all presence points
pred.vals <- raster::extract(pred.r, occ.xy)

# Minimum training presence threshold
or.min.threshold <- pred.r > min(pred.vals)

# convert pred vals to vector
pred.vect <- pred.vals[, 2]

mtp_threshold <- min(pred.vect)

sens_mtp_test <- mean(test.vals >= mtp_threshold)
sens_mtp_test 

# To impose the 10% presence threshold, first find how many records represent 10% of the records
n10 <- ceiling(length(pred.vect) * 0.1)

# Then impose the 10% threshold
or.10.threshold <- pred.r > sort(pred.vect)[n10]

or10_threshold_value <- sort(pred.vect)[n10]


# Plot to visualize
par(mfrow=c(2,2))
plot(pred.r, main='Continous')
points(eval.occs(m_kd_bias), pch=10, bg = eval.occs.grp(m_kd_bias))

plot(or.min.threshold, main='Min omission threshold')
points(eval.occs(m_kd_bias), pch=10, bg = eval.occs.grp(m_kd_bias))

plot(or.10.threshold, main='10% omission threshold')
points(eval.occs(m_kd_bias), pch=10, bg = eval.occs.grp(m_kd_bias))

# convert spatraster to df
or_min_df <- as.data.frame(or.min.threshold, xy = TRUE) 

or_min_df <- or_min_df %>% 
   rename(Suitability = opt.tune) %>% 
   mutate(Suitability = factor(Suitability))

occ.xy$occ_label <- "Training Occurrences"


or_min_map <- ggplot() +
  geom_raster(data = or_min_df, aes(x = x, y = y, fill = Suitability)) +
  scale_fill_manual(values = c("FALSE" = "lightgrey", "TRUE" = "palegreen3")) +
    geom_point(
    data = occ.xy,
    aes(x = x, y = y, color = occ_label),
    size = 1
  ) +
  scale_color_manual(
    name = "Points",
    values = c("Training Occurrences" = "black")
  ) +
  coord_sf(crs = "EPSG:4326", expand = FALSE) +
  labs(title = "Minimum Training Presence Threshold") +
  theme_bw() + #Clean theme
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.ticks = element_blank(), # Remove axis ticks,
    axis.text = element_blank(),  # Remove axis text (coordinate degrees)
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank()
  ) +
    annotation_scale(
    location   = "bl",
    width_hint = 0.25,
    text_cex   = 0.55,
    line_width = 0.6,
    pad_x      = unit(0.5, "npc"),
    pad_y      = unit(0.03, "npc")
  ) +
  annotation_north_arrow(
    location = "br",
    which_north = "true",
    height = unit(0.8, "cm"),
    width  = unit(0.8, "cm"),
    pad_x  = unit(0.02, "npc"),
    pad_y  = unit(0.02, "npc"),
  style = north_arrow_fancy_orienteering()
)

or_min_map

ggsave(
filename = file.path("Final_Results/GH_OR_min_suitability_map.tiff"),
plot = or_min_map,
device = "tiff",
dpi = 300,
width = 180,
height = 120,
units = "mm",
compression = "lzw"
)

```

```{r}
# extract suitability at bg locations
bg_train <- as.matrix(bg[, c("x","y")])
bg_vals <- terra::extract(pred.seq_kd_bias, bg_train)[,1]


e <- evaluate(
  p = train_vals,
  a = bg_vals
)

maxSSS_th <- threshold(e, "spec_sens")

sens_maxSSS_test <- mean(test.vals >= maxSSS_th)
sens_maxSSS_test
```

# Sensitivity table
```{r}
sens_thr_tbl <- tibble(
  Threshold = c("MTP", "OR10","MaxSSS"),
  `Threshold value` = c(mtp_threshold, th_or10, maxSSS_th),
  Sensitivity_value = c(sens_mtp_test, sens_OR10_test, sens_maxSSS_test
))

sens_thr_tbl

sens_thr_tbl_gt <- sens_thr_tbl %>% 
  mutate(`Sensitivity (%)` = Sensitivity_value * 100) %>% 
  dplyr::select(Threshold, `Threshold value`, `Sensitivity (%)`) %>% 
  gt() %>% 
  fmt_number(columns = c("Threshold value", "Sensitivity (%)"), decimals = 2)


sens_thr_tbl_gt

gtsave(sens_thr_tbl_gt, "Final_Results/ENMeval_Sensitivity_table.docx")
```


# OR 10% Threshold map
```{r}
# convert spatraster to df
or_10_df <- as.data.frame(or.10.threshold, xy = TRUE) 

or_10_df <- or_10_df %>% 
   rename(Suitability = opt.tune) %>% 
   mutate(Suitability = factor(Suitability))


or_10_map <- ggplot() +
  geom_raster(data = or_10_df, aes(x = x, y = y, fill = Suitability)) +
  scale_fill_manual(values = c("FALSE" = "lightgrey", "TRUE" = "palegreen3")) +
    geom_point(
    data = occ.xy,
    aes(x = x, y = y, color = occ_label),
    size = 1
  ) +
  scale_color_manual(
    name = "Points",
    values = c("Training Occurrences" = "black")
  ) +
  coord_sf(crs = "EPSG:4326", expand = FALSE) +
  labs(title = "10% OR Threshold") +
  theme_bw() + #Clean theme
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.ticks = element_blank(), # Remove axis ticks,
    axis.text = element_blank(),  # Remove axis text (coordinate degrees)
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank()
  ) +
    annotation_scale(
    location   = "bl",
    width_hint = 0.25,
    text_cex   = 0.55,
    line_width = 0.6,
    pad_x      = unit(0.5, "npc"),
    pad_y      = unit(0.03, "npc")
  ) +
  annotation_north_arrow(
    location = "br",
    which_north = "true",
    height = unit(0.8, "cm"),
    width  = unit(0.8, "cm"),
    pad_x  = unit(0.02, "npc"),
    pad_y  = unit(0.02, "npc"),
  style = north_arrow_fancy_orienteering()
)

or_10_map

ggsave(
filename = file.path("Final_Results/GH_OR_10_suitability_map.tiff"),
plot = or_10_map,
device = "tiff",
dpi = 300,
width = 180,
height = 120,
units = "mm",
compression = "lzw"
)
```


## Suitability Map based on MTP and OR10 thresholds
```{r}
# realign with template for bivariate mapping
pred.full <- terra::mask(terra::extend(pred.r, template), template)

# classify into 1, 2, 3
suit_class <- ifel(pred.full < mtp_threshold, 1,
                   ifel(pred.full < or10_threshold_value, 2, 3))

levels(suit_class) <- data.frame(
  ID = 1:3,
  class = c("Low", "Moderate", "High")
)

plot(suit_class)

ggplot() +
  theme_minimal(base_size = 14) +  # Set a minimal theme for the map
  geom_raster(data = suit_class, mapping = aes(x = x, y = y, fill = class), show.legend = FALSE) +
  geom_sf(data = gh_vect, fill = NA, color = "black", linewidth = 0.10)


writeRaster(suit_class, "Final_Results/env_suitability_thresholds.tif", overwrite = TRUE)
```


# Suitability - low, moderate, high
Low suitability: less than MTP thresholds
Moderate suitability: between MTP and 10p thresholds
High suitability: greater than 10p threshold

Categorization similar to https://www.mdpi.com/1660-4601/15/2/187#Materials_and_Methods>
```{r}
# convert spatraster to df
pred_df <- as.data.frame(pred.r, xy = TRUE)

pred_df <- pred_df %>%
   rename(Suitability = opt.tune) %>%
   mutate(Category = case_when(
     Suitability < mtp_threshold ~ "Low",
     Suitability >= mtp_threshold & Suitability <= or10_threshold_value ~ "Moderate",
     Suitability > or10_threshold_value ~ "High"
   ))

table(pred_df$Category)

color_categories <- c("Low" = "#D3D3D3", "Moderate" = "#7FBBD2", "High" = "#159ED0")

pred_df$Category <- factor(pred_df$Category, levels = c("Low", "Moderate", "High"))

suitability_category_map <- ggplot() +
  geom_raster(data = pred_df, aes(x = x, y = y, fill = Category)) +
  scale_fill_manual(values = color_categories, name = "Suitability Category") +
  # geom_point(data = occ.xy,
  #            aes(x = x, y = y),
  #                color = "black",
  #                size = 1) +
  coord_sf(crs = "EPSG:4326", expand = FALSE) +
  theme_bw() + #Clean theme
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.ticks = element_blank(), # Remove axis ticks,
    axis.text = element_blank(),  # Remove axis text (coordinate degrees)
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank()
  ) 

suitability_category_map
```

```{r}
ggsave(
filename = file.path("Final_Results/environmental_suitability_categories.tiff"),
plot = suitability_category_map,
device = "tiff",
dpi = 300,
width = 180,
height = 120,
units = "mm",
compression = "lzw"
)
```

